Rates of true polar wander in convecting planets

### 2014-02-20 tpw_rate.tex
Add extra option for more citation options
Add some appendix text about MacCullagh

%\documentclass[onecolumn]{gji}
\documentclass{gji}
\usepackage{timet}
\usepackage{amsmath}

@@

@@

\begin{equation}
\begin{aligned}
 \rho_0 \left[ \frac{D \bf v}{D t} + \frac{d \bf \Omega}{dt} \times {\bf r} +  2 {\bf \Omega} \times {\bf v} + {\bf \Omega \times \Omega \times r} \right] \\ = - \nabla P + \nabla \cdot \left( \eta \nabla_s {\bf v} \right) + \rho {\bf g}
\label{navier_stokes}
\end{aligned}
\end{equation}

@@

along with the equation of state

@@

Dimensional analysis of this system (c.f. \cite{barenblatt1996scaling}) requires six nondimensional numbers to characterize it.

@@

>>>>

%\documentclass[extra,onecolumn]{gji}
\documentclass[extra]{gji}
\usepackage{timet}
\usepackage{amsmath}

@@

%in case I want to add detail to any derivation
\newif\ifdetail
\detailfalse

@@

\begin{equation}
\begin{aligned}
 \rho_0 \left[ \frac{D \bf v}{D t} + \frac{d \bf \Omega}{dt} \times {\bf r} +  2 {\bf \Omega} \times {\bf v} + {\bf \Omega \times \Omega \times r} \right] \\ = - \nabla P + \nabla \cdot \left( \eta \varepsilon ({\bf v}) \right) + \rho {\bf g}
\label{navier_stokes}
\end{aligned}
\end{equation}

@@

 where the parameters are defined in Table~\ref{parameters} and $\varepsilon({\bf v}) = 1/2 \; ( \partial v_i / \partial x_j + \partial v_j / \partial x_i )$ corresponds to the symmetrized velocity gradient.  In addition we use the simple equation of state

@@

Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires six nondimensional numbers to characterize it.

@@

\appendix

\section{Degree-two moments}

We may explicitly draw a connection between the moment of inertia of a rotating object and it's degree-two density structure.  The moment of inertia tensor may be written in index notation
\begin{equation}
I_{ij} = \int_V \rho \left( r_q r_q \delta_{ij} - r_i r_j \right) d^3 {\bf r}
\label{inertia}
\end{equation}

where $\bf r$ is the Eulerian coordinate, $\rho$ is the density, and $V$ is the volume of the material.  The degree-two density structure may be expressed as a quadrupole-moment tensor (cf. \citet{jackson1998classical}):

\begin{equation}
Q_{ij} = \int_V \rho \left( 3 r_i r_j - r_q r_q \delta_{ij} \right) d^3 {\bf r}
\end{equation}

The commutator of these tensors, $\bf QI - IQ$, is zero, indicating that they may be simultaneously diagonalized.

\ifdetail
We may demonstrate that the commutator is zero by using the definition of the commutator and expanding:
\begin{equation} 
\begin{aligned}
{\bf IQ-QI} = &I_{ik} Q_{kj} - Q_{ik} I_{kj} \\
 = &\int_V \rho \left( r_q r_q \delta_{ik} - r_i r_k \right) d^3 {\bf r} \int_V \rho \left( 3 r_k r_j - r_q r_q \delta_{kj} \right) d^3 {\bf r} \\
 &- \int_V \rho \left( r_q r_q \delta_{kj} - r_k r_j \right) d^3 {\bf r} \int_V \rho \left( 3 r_i r_k - r_q r_q \delta_{ik} \right) d^3 {\bf r} \\
 = &3 \delta_{ik} \int_V \rho r_q r_q d^3 {\bf r} \int_V \rho r_k r_j d^3 {\bf r} \\
 &-\delta_{kj} \delta_{ik} \int_V \rho r_q r_q d^3 {\bf r} \int_V \rho r_q r_q d^3 {\bf r} \\
 &-3 \int_V \rho r_i r_k d^3 {\bf r} \int_V \rho r_k r_j d^3 {\bf r} \\
 &+\delta_{kj} \int_V \rho r_i r_k d^3 {\bf r} \int_V \rho r_q r_q d^3 {\bf r} \\
 - &3 \delta_{kj} \int_V \rho r_q r_q d^3 {\bf r} \int_V \rho r_i r_k d^3 {\bf r} \\
 &+\delta_{ik} \delta_{kj} \int_V \rho r_q r_q d^3 {\bf r} \int_V \rho r_q r_q d^3 {\bf r} \\
 &+3 \int_V \rho r_k r_j d^3 {\bf r} \int_V \rho r_i r_k d^3 {\bf r} \\
 &-\delta_{ik} \int_V \rho r_k r_j d^3 {\bf r} \int_V \rho r_q r_q d^3 {\bf r} \\
 = \;&3 \int_V \rho r_q r_q d^3 {\bf r} \int_V \rho r_i r_j d^3 {\bf r} \\
 &+ \int_V \rho r_i r_j d^3 {\bf r} \int_V \rho r_q r_q d^3 {\bf r} \\
 - &3 \int_V \rho r_q r_q d^3 {\bf r} \int_V \rho r_i r_j d^3 {\bf r} \\
 &- \int_V \rho r_i r_j d^3 {\bf r} \int_V \rho r_q r_q d^3 {\bf r} \\
 & = 0
\end{aligned}
\end{equation}
\else
\fi

We may therefore go into the principle coordinate system of both tensors to find
\begin{equation}
{\bf I} = \begin{bmatrix}
\int_V \rho (y^2+z^z)\\
\int_V \rho (x^2+z^2) \\
\int_V \rho (x^2+y^2) 
\end{bmatrix}^T {\bf 1}
\end{equation}

\begin{equation}
{\bf Q} = \begin{bmatrix}
\int_V \rho (2 x^2 - y^2 - z^z)\\
\int_V \rho (2 y^2 - x^2 -z^2)\\
\int_V \rho (2 z^2 - x^2 - y^2) 
\end{bmatrix}^T {\bf 1}
\end{equation}

where $\bf 1$ is the identity matrix.  If we let the diagonal elements of $\bf I$ be $\lambda_1, \lambda_2, \lambda_3$, we may rewrite $\bf Q$ as

\begin{equation}
{\bf Q} = \begin{bmatrix}
(\lambda_2-\lambda_1) + (\lambda_3-\lambda_1)\\
(\lambda_1-\lambda_2) + (\lambda_3-\lambda_2)\\
(\lambda_1-\lambda_3) + (\lambda_2-\lambda_3)\\
\end{bmatrix}^T {\bf 1}
\label{maccullagh}
\end{equation}

which shows the connection between the degree-two component of the density field and the principle moments of inertia.  This may be seen as a form of MacCullagh's formula \citep{stacey1977physics}
